---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rSDP

<!-- badges: start -->
`r lifecycle::badge('experimental')`
<!-- badges: end -->

The rSDP package provides a simplified interface for discovering, querying, and subsetting data products that are incorporated into the RMBL Spatial Data Platform. The RMBL SDP provides a set of curated, high-resolution, and high-fidelity geospatial datasets for a set of domains in Western Colorado (USA) in the vicinity of [Rocky Mountain Biological Laboratory](https://rmbl.org). For more information about the RMBL SDP [see here](https://www.rmbl.org/scientists/resources/spatial-data-platform/).

SDP data products are provided as geospatial raster datasets in [cloud-optimized Geotiff]() (COG) format. The rSDP package provides functions to access these datasets in cloud storage (Amazon S3) without downloading.

## Installation

You can install the development version of rSDP from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("rmbl-sdp/rSDP")
```

## Discovering SDP Data and Metadata

The package provides functions `sdp_get_catalog()`, and `sdp_get_metadata()` that download information about what datasets are currently available and what their spatial attributes are.

```{r example}
library(rSDP)
## Gets entries for vegetation data products in the Upper Gunnison (UG) domain.
sdp_cat <- sdp_get_catalog(domains="UG", 
                           types="Vegetation",
                           deprecated=FALSE,
                           return_stac=FALSE)
sdp_cat[,1:5]
```

```{r example2}
## Grabs detailed metadata for a specific dataset.
item_meta <- sdp_get_metadata(catalog_id="R1D001",return_list=TRUE)

## Prints the detailed description.
item_description <- item_meta$qgis$abstract[[1]]
print(item_description)
```
